---
title: "Counting Points In a Buffer"
author: "Susannah Ripley"
date: "20/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The City of Toronto has an amazing database of all the trees on road allowances. This database has over half a million trees with species and trunk diameter. If you're a Toronto resident, you can look up trees near you [here](http://trees.jordanteichmann.ca/).

I am interested in seeing if the abundance of these urban trees might be a useful predictor in a land-use regression model.  

I have air pollutants measured at 60 sites across the city. My goal is to count the number of trees within a 500 m radius of each of these points. I then want to assign this count value to a variable in my dataset.  

We will need the **SimpleFeatures (sf)** package as well as the **tidyverse** package.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(sf)
library(rgdal)
```

## Load my dataset of points

My dataset consists of geographic coordinates as latitude and longitude, as well as a pollutant concentration measured at each location.

```{r cars}
df <- read.csv("./data/points.csv")
head(df)
```

## Make 500 m buffers around the points

We currently have points as a dataframe. We will use the st_as_sf function to convert this dataframe to a spatial object. 

```{r}
pointsto <-  st_as_sf(df, coords = c("longitude", "latitude"), crs="NAD83") 
str(pointsto)
```

Currently the coordinates are expressed as latitude and longitude. To use the **st_buffer** function that we will use to create a buffer around each point, it is easiest if we have a coordinate system that has meters as units of measurements (like UTM). 

Toronto is in UTM zone 17N, which is represented by EPSG 32617. We will transform the points object into this projection.

```{r}
pointsto <- st_transform(pointsto, crs=32617)
```

Now we can create the buffers. The function **st_buffer** takes 2 arguments: the object around which buffers are to be made, and the buffer distance. I chose a buffer of radius 500 m but you can specify whatever distance your heart desires. 

```{r}
buffsto500 <- st_buffer(pointsto, dist=500)
```

Let's map this to make sure everything looks good.

```{r}
ggplot() +
  geom_sf(data=buffsto500, color="pink") +
  geom_sf(data=pointsto) 
```

## Add Toronto Street Trees Data

Next up we will download the Street Trees data from the City of Toronto's Open Data Portal. 

```{r}
URL <- "https://ckan0.cf.opendata.inter.prod-toronto.ca/download_resource/c1229af1-8ab6-4c71-b131-8be12da59c8e"
download.file(url=URL, destfile="./data/TO street trees.zip", mode="wb")

trees <- unzip(zipfile="./data/TO street trees.zip", exdir = "./data/trees")
trees <- readOGR("./data/trees")
# Make it into a spatial SimpleFeatures object
trees <- st_as_sf(trees, crs="WGS84")
# Reproject to UTM
trees <- st_transform(trees, crs=32617)
str(trees)
```

You can see that this dataset has 1 line per tree. Let's plot it to see where the trees are, and because it looks cool.

```{r}
ggplot() + geom_sf(data=trees, color="darkgreen", alpha=0.05, shape=".")
```

Very cool.

## Count number of rows of trees dataset within the buffer

We will use the function **st_intersects** to determine how many elements of the *trees* dataset fall within or on the border of the 500 m buffer around each point. Then we will count the length of each intersection and create a new variable representing this count.

```{r}
df$trees_count_500m <- lengths(st_intersects(buffsto500, trees))
buffsto500$trees_count_500m <- lengths(st_intersects(buffsto500, trees))
```

Let's do a histogram to examine the frequency of different tree counts.  

```{r}
ggplot() +
  geom_histogram(data=df, aes(x=trees_count_500m), fill="darkgreen") +
  labs(x="Count of trees in 500 m buffer", y="Number of points", main="Frequency of tree counts in 500 m buffers") +
  theme_minimal()
```

Finally, we will map the count of trees within each buffer to visualize the data.

```{r}
ggplot() +
  geom_sf(data=buffsto500, aes(fill=trees_count_500m)) +
  scale_fill_distiller(palette="Greens", direction = 1) +
  geom_sf(data=pointsto) +
  labs(main="Toronto Tree Count In 500 m Buffer") + 
  labs(fill="Trees in 500 m buffer")
```